---
title: 'Fire/CC effects on species based on estimated coefficients'
author: "Peter Solymos"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: word_document
---

# Introduction

The goal is to use coefficients from our all-in-one analysis (as posted on the species website)
and generalize how fire vs. cutblock affect species and taxa.
This hopefully simple analysis can feed into the ACTWS presentation.


# Methods

I compiled coefficients for White Spruce, Pine, Mixedwood, and Deciduous forest stands from
all taxa (except for mammals).

I averaged coefficients in upland forest, lowland forest and open habitats.
I defined upland forest species as having a >0.5 upland forest score
based on sum-standardized mean abundances amon the 3 broad land cover types.
Note that this eliminates some generalist species which might well live in upland forests.

I further defined early seral upland forest species that showed 
highest relative abundance in the 0-40 years post harvest/fire classes.

I used relative abundances within the 4 stand types 
(for each species standardized by maximum across all upland stand types to reflect differences across species)
to cluster age classes based on species' relative abundances.
I classified age classes based on natural stands alone, or natural and harvested stands combined.
I used Bray-Curtis dissimilarity and Ward's agglomerative method for clustering.

Fire vs. forest harvest associations were quantified based on the 0-9 yr age classes.
The rescaled (-1, 1 interval) ratio of harvest / fire was used as an affinity index.
I also compared the 90\% confidence intervals around the 0-9 yr age class coefficients
and tested if the intervals overlaped, and tested the direction when those did not overlap.
This is not an exact test, but takes into account uncertainty, unlike the simple affinity index above.

The analyses were repeated for each stand type.

# Results

```{r echo=FALSE,message=FALSE,warning=FALSE}
library(vegan)
library(ape)
library(mefa4)
library(intrval)
library(knitr)
load("~/Dropbox/abmi/fire-cc-all.Rdata")
col1 <- colorRampPalette(c("red", "darkgreen"))(9)
col2 <- c(col1, colorRampPalette(c("blue", col1[5]))(5))
Age <- as.integer(gsub("[[:alpha:]]", "", colnames(all[[1]][[2]])))

data_fun <- function(taxon, ftype, cc=FALSE) {
    tmp <- all[[taxon]][[paste0(ftype, "_fire")]]
    if (cc)
        tmp <- cbind(tmp, all[[taxon]][[paste0(ftype, "_cc")]])
    Max <- sapply(all[[taxon]][c("conif_fire", "decid_fire", "pine_fire", "mixed_fire",
        "conif_cc", "decid_cc", "pine_cc", "mixed_cc")], function(z) apply(z, 1, max, na.rm=TRUE))
    Max <- apply(Max, 1, max)
    tmp <- tmp / Max
    tmp
}

## Upland species
upfor <- c("WhiteSpruce0", "WhiteSpruce10", "WhiteSpruce20",
    "WhiteSpruce40", "WhiteSpruce60", "WhiteSpruce80", "WhiteSpruce100",
    "WhiteSpruce120", "WhiteSpruce140", "Pine0", "Pine10", "Pine20",
    "Pine40", "Pine60", "Pine80", "Pine100", "Pine120", "Pine140",
    "Deciduous0", "Deciduous10", "Deciduous20", "Deciduous40", "Deciduous60",
    "Deciduous80", "Deciduous100", "Deciduous120", "Deciduous140",
    "Mixedwood0", "Mixedwood10", "Mixedwood20", "Mixedwood40", "Mixedwood60",
    "Mixedwood80", "Mixedwood100", "Mixedwood120", "Mixedwood140")
lowfor <- c("BlackSpruce0", "BlackSpruce10", "BlackSpruce20", "BlackSpruce40",
    "BlackSpruce60", "BlackSpruce80", "BlackSpruce100", "BlackSpruce120",
    "BlackSpruce140", "Larch0", "Larch10", "Larch20", "Larch40",
    "Larch60", "Larch80", "Larch100", "Larch120", "Larch140")
open <- c("Swamp", "WetGrass", "WetShrub", "Shrub", "GrassHerb")
lt <- list()
for (i in names(all)) {
    mm <- data.frame(upfor = rowMeans(all[[i]]$hab[,colnames(all[[i]]$hab) %in% upfor], na.rm=TRUE),
        lowfor = rowMeans(all[[i]]$hab[,colnames(all[[i]]$hab) %in% lowfor], na.rm=TRUE),
        open = rowMeans(all[[i]]$hab[,colnames(all[[i]]$hab) %in% open], na.rm=TRUE))
    lt[[i]] <- data.frame(taxon=i, all[[i]]$spp, mm)
}
lt <- do.call(rbind, lt)
mmm <- lt[,c("upfor","lowfor","open")]
mmm <- mmm / rowSums(mmm)
#mmm[mmm[,"open"] > 0.5,]
#mmm[mmm[,"lowfor"] > 0.5,]
#mmm[mmm[,"open"] <= 0.5 & mmm[,"lowfor"] <= 0.5,]
lt$pref <- factor("general", levels=c("upfor", "lowfor", "open", "general"))
lt$pref[mmm[,"upfor"] > 0.5] <- "upfor"
lt$pref[mmm[,"lowfor"] > 0.5] <- "lowfor"
lt$pref[mmm[,"open"] > 0.5] <- "open"

## CI based comparison
cio <- list()
cig <- list()
for (i in names(all)) {
    tmp <- all[[i]]$hab[,c("WhiteSpruce0.LCL","WhiteSpruce0.UCL",
        "WhiteSpruceCC0.LCL","WhiteSpruceCC0.UCL",
        "Pine0.LCL","Pine0.UCL",
        "PineCC0.LCL","PineCC0.UCL",
        "Mixedwood0.LCL","Mixedwood0.UCL",
        "MixedwoodCC0.LCL","MixedwoodCC0.UCL",
        "Deciduous0.LCL","Deciduous0.UCL",
        "DeciduousCC0.LCL","DeciduousCC0.UCL")]
    cio[[i]] <- data.frame(
        conif= tmp[,1:2] %)o(% tmp[,3:4],
        pine=tmp[,5:6] %)o(% tmp[,7:8],
        mixed=tmp[,9:10] %)o(% tmp[,11:12],
        decid=tmp[,13:14] %)o(% tmp[,15:16])
    cig[[i]] <- data.frame(
        conif= tmp[,1:2] %[o>]% tmp[,3:4],
        pine=tmp[,5:6] %[o>]% tmp[,7:8],
        mixed=tmp[,9:10] %[o>]% tmp[,11:12],
        decid=tmp[,13:14] %[o>]% tmp[,15:16])
}
cio <- do.call(rbind, cio)
cig <- do.call(rbind, cig)

lt <- data.frame(lt, out=cio, fgtcc=cig)

lt$ci.conif <- factor("f=c", levels=c("f>c", "f=c", "f<c"))
lt$ci.conif[!lt$out.conif] <- "f=c"
lt$ci.conif[lt$out.conif & lt$fgtcc.conif] <- "f>c"
lt$ci.conif[lt$out.conif & !lt$fgtcc.conif] <- "f<c"

lt$ci.pine <- factor("f=c", levels=c("f>c", "f=c", "f<c"))
lt$ci.pine[!lt$out.pine] <- "f=c"
lt$ci.pine[lt$out.pine & lt$fgtcc.pine] <- "f>c"
lt$ci.pine[lt$out.pine & !lt$fgtcc.pine] <- "f<c"

lt$ci.mixed <- factor("f=c", levels=c("f>c", "f=c", "f<c"))
lt$ci.mixed[!lt$out.mixed] <- "f=c"
lt$ci.mixed[lt$out.mixed & lt$fgtcc.mixed] <- "f>c"
lt$ci.mixed[lt$out.mixed & !lt$fgtcc.mixed] <- "f<c"

lt$ci.decid <- factor("f=c", levels=c("f>c", "f=c", "f<c"))
lt$ci.decid[!lt$out.decid] <- "f=c"
lt$ci.decid[lt$out.decid & lt$fgtcc.decid] <- "f>c"
lt$ci.decid[lt$out.decid & !lt$fgtcc.decid] <- "f<c"
```

## Pine

```{r echo=FALSE}
kable(as.matrix(ftable(lt[lt$pref=="upfor",c("taxon", "ci.pine")])))
```

```{r echo=FALSE,fig.height=4,fig.width=8}
ftype <- "pine"
cc <- FALSE
tmp1 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
cc <- TRUE
tmp2 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
tmp1 <- tmp1[lt$pref=="upfor",]
tmp2 <- tmp2[lt$pref=="upfor",]

op <- par(mfrow=c(1,2))
plot(as.phylo(hclust(vegdist(t(tmp1)), method="ward.D2"), main="All species"),
    tip.color=col1, font=2, main=ftype)
plot(as.phylo(hclust(vegdist(t(tmp2)), method="ward.D2"), main="All species"),
    tip.color=col2, font=2)
par(op)
```

```{r echo=FALSE,fig.height=7,fig.width=7}
dat <- data.frame(taxon=lt$taxon[lt$pref=="upfor"], rr = tanh(0.5 * log(tmp2[,10] / tmp2[,1])))
mx <- find_max(tmp2)
ii <- as.integer(mx$index) %in% c(1,2,3, 10,11,12)
op <- par(mfrow=c(2,1))
boxplot(rr ~ taxon, dat, main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland species")
abline(h=0, col=4, lwd=2)
boxplot(rr ~ taxon, dat[ii,], main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland early-seral species")
abline(h=0, col=4, lwd=2)
par(op)
```



## White Spruce

```{r echo=FALSE}
kable(as.matrix(ftable(lt[lt$pref=="upfor",c("taxon", "ci.conif")])))
```

```{r echo=FALSE,fig.height=4,fig.width=8}
ftype <- "conif"
cc <- FALSE
tmp1 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
cc <- TRUE
tmp2 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
tmp1 <- tmp1[lt$pref=="upfor",]
tmp2 <- tmp2[lt$pref=="upfor",]

op <- par(mfrow=c(1,2))
plot(as.phylo(hclust(vegdist(t(tmp1)), method="ward.D2"), main="All species"),
    tip.color=col1, font=2, main=ftype)
plot(as.phylo(hclust(vegdist(t(tmp2)), method="ward.D2"), main="All species"),
    tip.color=col2, font=2)
par(op)
```

```{r echo=FALSE,fig.height=7,fig.width=7}
dat <- data.frame(taxon=lt$taxon[lt$pref=="upfor"], rr = tanh(0.5 * log(tmp2[,10] / tmp2[,1])))
mx <- find_max(tmp2)
ii <- as.integer(mx$index) %in% c(1,2,3, 10,11,12)
op <- par(mfrow=c(2,1))
boxplot(rr ~ taxon, dat, main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland species")
abline(h=0, col=4, lwd=2)
boxplot(rr ~ taxon, dat[ii,], main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland early-seral species")
abline(h=0, col=4, lwd=2)
par(op)
```

## Mixedwood

```{r echo=FALSE}
kable(as.matrix(ftable(lt[lt$pref=="upfor",c("taxon", "ci.mixed")])))
```

```{r echo=FALSE,fig.height=4,fig.width=8}
ftype <- "mixed"
cc <- FALSE
tmp1 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
cc <- TRUE
tmp2 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
tmp1 <- tmp1[lt$pref=="upfor",]
tmp2 <- tmp2[lt$pref=="upfor",]

op <- par(mfrow=c(1,2))
plot(as.phylo(hclust(vegdist(t(tmp1)), method="ward.D2"), main="All species"),
    tip.color=col1, font=2, main=ftype)
plot(as.phylo(hclust(vegdist(t(tmp2)), method="ward.D2"), main="All species"),
    tip.color=col2, font=2)
par(op)
```

```{r echo=FALSE,fig.height=7,fig.width=7}
dat <- data.frame(taxon=lt$taxon[lt$pref=="upfor"], rr = tanh(0.5 * log(tmp2[,10] / tmp2[,1])))
mx <- find_max(tmp2)
ii <- as.integer(mx$index) %in% c(1,2,3, 10,11,12)
op <- par(mfrow=c(2,1))
boxplot(rr ~ taxon, dat, main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland species")
abline(h=0, col=4, lwd=2)
boxplot(rr ~ taxon, dat[ii,], main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland early-seral species")
abline(h=0, col=4, lwd=2)
par(op)
```


## Deciduous

```{r echo=FALSE}
kable(as.matrix(ftable(lt[lt$pref=="upfor",c("taxon", "ci.decid")])))
```

```{r echo=FALSE,fig.height=4,fig.width=8}
ftype <- "decid"
cc <- FALSE
tmp1 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
cc <- TRUE
tmp2 <- rbind(data_fun("birds", ftype, cc), data_fun("lichens", ftype, cc),
    data_fun("mites", ftype, cc), data_fun("mosses", ftype, cc),
    data_fun("vplants", ftype, cc))
tmp1 <- tmp1[lt$pref=="upfor",]
tmp2 <- tmp2[lt$pref=="upfor",]

op <- par(mfrow=c(1,2))
plot(as.phylo(hclust(vegdist(t(tmp1)), method="ward.D2"), main="All species"),
    tip.color=col1, font=2, main=ftype)
plot(as.phylo(hclust(vegdist(t(tmp2)), method="ward.D2"), main="All species"),
    tip.color=col2, font=2)
par(op)
```

```{r echo=FALSE,fig.height=7,fig.width=7}
dat <- data.frame(taxon=lt$taxon[lt$pref=="upfor"], rr = tanh(0.5 * log(tmp2[,10] / tmp2[,1])))
mx <- find_max(tmp2)
ii <- as.integer(mx$index) %in% c(1,2,3, 10,11,12)
op <- par(mfrow=c(2,1))
boxplot(rr ~ taxon, dat, main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland species")
abline(h=0, col=4, lwd=2)
boxplot(rr ~ taxon, dat[ii,], main=ftype, col="lightgreen",
    ylab="Cutblock Affinity", sub="Upland early-seral species")
abline(h=0, col=4, lwd=2)
par(op)
```

