---
title: "Bird population size estimates in BCR 6 Alberta"
author: "Peter Solymos (solymos@ualberta.ca)"
output:
  pdf_document: default
  word_document: default
---

```{r load_data,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}
library(mefa4)

ROOT <- "e:/peter/AB_data_v2016"

e <- new.env()
load(file.path(ROOT, "out", "birds", "data", "data-wrsi.Rdata"), envir=e)
TAX <- droplevels(e$TAX)
TAX$Fn <- droplevels(TAX$English_Name)
levels(TAX$Fn) <- nameAlnum(levels(TAX$Fn), capitalize="mixed", collapse="")
rm(e)

e <- new.env()
load(file.path(ROOT, "out", "birds", "data", "data-josmshf.Rdata"), envir=e)
xn <- e$DAT
mods <- e$mods
OFF <- e$OFF
yy <- e$YY
BB <- e$BB
rm(e)

fln <- list.files(file.path(ROOT, "out", "birds", "results", "josmshf"))
fln <- sub("birds_abmi-josmshf_", "", fln)
fln <- sub(".Rdata", "", fln)

SPP <- fln
tax <- droplevels(TAX[SPP,])

tv0 <- read.csv("~/repos/abmianalytics/lookup/lookup-veg-hf-age.csv")
tv0$Sector2 <- factor(ifelse(is.na(tv0$Sector), "NATIVE", as.character(tv0$Sector)),
    c("NATIVE", "Agriculture", "Energy", "Forestry", "Misc", "RuralUrban", "Transportation"))
tv0$HAB <- paste0(tv0$Type, ifelse(tv0$AGE %in% c("5", "6", "7", "8", "9"), "O", ""))
tv0$HAB[tv0$Combined %in% c("RoadTrailVegetated", "RoadVegetatedVerge",
    "RailVegetatedVerge")] <- "Verges"

## PIF table
pif <- read.csv("~/Dropbox/bam/PIF-AB/popBCR-6AB_v2_22-May-2013.csv")
mefa4::compare_sets(tax$English_Name, pif$Common_Name)
setdiff(tax$English_Name, pif$Common_Name)
pif <- pif[match(tax$English_Name, pif$Common_Name),]
rownames(pif) <- rownames(tax)

AUC <- read.csv("~/Dropbox/bam/PIF-AB/results/AUC.csv")
rownames(AUC) <- AUC$spp
AUC$spp <- NULL
AUC <- AUC[rownames(tax),]

if (FALSE) {
NN <- read.csv("~/Dropbox/bam/PIF-AB/results/PopSize567.csv")
rownames(NN) <- NN$spp
NN$spp <- NULL
NN <- NN[rownames(tax),]

load("e:/peter/josm/2017/stage7/predB/predictionsCI.Rdata")
rm(PREDSCI0)
PREDSCI <- PREDSCI[,rownames(tax),] / 10^6
N7B <- apply(PREDSCI, 2, colSums)
N7CI <- t(apply(N7B, 2, quantile, c(0.5, 0.025, 0.975)))
N7mean <- colMeans(N7B)
}

load("e:/peter/AB_data_v2016/out/birds/data/mean-qpad-estimates.Rdata")
qpad_vals <- qpad_vals[rownames(tax),]

## roadside_bias, rai_pred, rai_data
load("e:/peter/josm/2017/roadside_avoidance.Rdata")
rsb <- t(sapply(roadside_bias, function(z) z[,1]))

## AvegH, Nhab
load("e:/peter/josm/2017/stage7/predB/predictions_HAB.Rdata")

## bootstrap averaged pop size estimate
b_fun <- function(h) {
    N7tb <- c(Mean=mean(colSums(h) / 10^6),
        quantile(colSums(h) / 10^6, c(0.5, 0.025, 0.975)))
    N7hb <- t(apply(h, 1, function(z)
        c(Mean=mean(z / 10^6), quantile(z / 10^6, c(0.5, 0.025, 0.975)))))
    N7b <- rbind(N7hb, TOTAL=N7tb)
    N7b
}
NestAll <- lapply(Nhab, b_fun)
NestTot <- t(sapply(NestAll, function(z) z["TOTAL",]))

library(mefa4)
library(rgdal)
library(rgeos)
library(sp)
library(gstat)
library(raster)
load(file.path("e:/peter/AB_data_v2016", "out", "kgrid", "kgrid_table.Rdata"))

r <- raster(file.path("~/Dropbox/courses/st-johns-2017",
    "data", "ABrasters", "dem.asc"))
slope <- terrain(r, opt="slope")
aspect <- terrain(r, opt="aspect")
hill <- hillShade(slope, aspect, 40, 270)

od <- setwd("e:/peter/AB_data_v2017/data/raw/xy/bcr/")
BCR <- readOGR(".", "BCR_Terrestrial_master") # rgdal
BCR <- spTransform(BCR, proj4string(r))
BCR <- gSimplify(BCR, tol=500, topologyPreserve=TRUE)
setwd(od)

od <- setwd("~/Dropbox/courses/st-johns-2017/data/NatRegAB")
AB <- readOGR(".", "Natural_Regions_Subregions_of_Alberta") # rgdal
AB <- spTransform(AB, proj4string(r))
AB <- gUnaryUnion(AB, rep(1, nrow(AB))) # province
AB <- gSimplify(AB, tol=500, topologyPreserve=TRUE)
setwd(od)

BCR2AB <- gIntersection(AB, BCR, byid=TRUE)

For <- c("Decid", "Mixwood", "Pine", "Conif", "BSpr", "Larch")
xn$HAB <- paste0(xn$hab1, ifelse(xn$hab1 %in% For & xn$wtAge*200 >= 80, "O", ""))
compare_sets(colnames(AvegH), xn$HAB)
setdiff(colnames(AvegH), xn$HAB)
xnss <- nonDuplicated(xn, SS, TRUE)
xy <- xnss
coordinates(xy) <- ~ POINT_X + POINT_Y
proj4string(xy) <-
    CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
xy <- spTransform(xy, proj4string(r))
xy2BCR <- over(xy, BCR)

tmp <- xnss[!is.na(xy2BCR) & xy2BCR==24,]
tmp <- tmp[!(tmp$PCODE != "BBSAB" & tmp$ROAD01 > 0),]
tab <- table(tmp$HAB, tmp$ROAD01)

#Ahab <- tab[,"0"] / sum(tab[,"0"])
Ahab <- colSums(AvegH[,rownames(tab)]) / sum(AvegH[,rownames(tab)])
Whab <- tab[,"1"] / sum(tab[,"1"])
AWhab <- data.frame(Ahab, Whab)
NAM <- names(Ahab)

h_fun <- function(h) {
    NN <- h[NAM, "Mean"] * 10^6 # back to individuals
    DD <- NN / colSums(AvegH)[NAM] # density: males / ha
    sum(DD * Whab) / sum(DD * Ahab)
}
H <- sapply(NestAll, h_fun)

## taxonomy etc ---
pop <- tax[,c("Species_ID", "English_Name", "Scientific_Name", "Spp")]
## pop size estimates ---
#pop$Npix <- NestTot[,"Mean"] # M males
pop$Npix <- NestTot[,"50%"]
pop$NpixLo <- NestTot[,"2.5%"]
pop$NpixHi <- NestTot[,"97.5%"]
pop$Npif <- (pif$Population_Estimate_unrounded / pif$Pair_Adjust) / 10^6 # M males
## roadside related metrics
pop$Y1 <- rsb[,"Lam1"]
pop$Y0 <- rsb[,"Lam0"]
## Tadj and EDR/MDD ---
pop$TimeAdj <- pif$Time_Adjust
#pop$p3 <- 1-exp(-3 * qpad_vals$phi0)
pop$p3 <- 1-exp(-3 * qpad_vals$phi)
pop$MDD <- pif$Detection_Distance_m
#pop$EDR <- qpad_vals$tau0 * 100
pop$EDR <- qpad_vals$tau * 100
## QAQC ---
pop$AUCin <- AUC$auc7i
pop$AUCout <- AUC$auc7o
pop$k7 <- AUC$k7
pop$DataQ <- pif$Data_Quality_Rating
pop$BbsVar <- pif$BBS_Variance_Rating
pop$SpSamp <- pif$Species_Sample_Rating
pop$H <- H
## Deltas ---
pop$DeltaObs <- log(pop$Npix / pop$Npif)
pop$DeltaR <- log(pop$Y0/pop$Y1)
pop$DeltaT <- log((1/pop$p3)/pop$TimeAdj)
pop$DeltaA <- log((1/pop$EDR^2) / (1/pop$MDD^2))
pop$DeltaH <- log(1/H) # we take inverse because H=1 is the PIF setup
pop$DeltaExp <- pop$DeltaR + pop$DeltaT + pop$DeltaA + pop$DeltaH
pop$epsilon <- pop$DeltaObs - pop$DeltaExp
## subset ---
pop <- droplevels(pop[rowSums(is.na(pop))==0,])
pop <- pop[sort(rownames(pop)),]
```

## Introduction

Existing populations size estimates by
Partners in Flight (PIF) provide a set of
population size estimates for bird species in
Bird Conservation Region (BCR) 6 in Alberta.
For any given species, we refer to this estimate
as the 'PIF estimate', $\hat{N}^{(PIF)}$.

The estimator for the PIF approach is:

$$ \frac{\hat{N}^{(PIF)}}{H} = E[Y_{1}] \times P \times T \times A \times \frac{1}{\pi MDD^2} $$

where $E[Y_{1}]$ is the stop level expected count from roadside BBS data (the original
description used route level averages and a factor or 1/50 to account for
differences in sampling area),
$P$ is the pair adjustment,
$T$ is the time adjustment,
$A$ is the area of the geographic region,
$MDD$ is maximum detection distance
(making sure that unit for $MDD$ is compatible to the unit of area),
and $1/H$ is adjustment for habitat sampling bias
due to applying a sample based estimator for the whole geographic region.
The habitat sampling bias $1/H$ is on the left hand side because current
PIF estimates do not account for that (tables contain a default value of $H=1$).

Another set of estimates of current population sizes in Alberta BCR 6
are based on a model based prediction approach
using province wide vegetation and human footprint maps.
We refer to these estimates for any given species
as the 'pixel based estimates', $\hat{N}^{(PIX)}$.

$$ \hat{N}^{(PIX)} = E[Y_{0}] \times P \times \frac{1}{p_3} \times A \times \frac{1}{\pi EDR^2} $$

where $E[Y_{0}]$ is the point level expected value of a count from a combination of
on- and off-road data sources after accounting for roadside bias,
$P$ is the pair adjustment,
$1/p_3$ is the time adjustment based on removal sampling based availability and 3 minutes time interval,
$A$ is the area of the geographic region,
$EDR$ is distance sampling based effective detection radius
(making sure that unit for $EDR$ is compatible to the unit of area),
There is no adjustment for habitat sampling bias because we used 
model based extrapolation to the whole geographic region to avoid habitat 
related biases in population size estimation.

We used the Pair adjustment to calculate number of singinsg individuals
for both estimators, thus effectively setting $P=1$.
We compare the two estimators and the different adjustments as ratios
(pair adjustment and the area of the geographic area calcels out):

$$ \frac{\hat{N}^{(PIX)}/P}{\hat{N}^{(PIF)}/(HP)} = \frac{E[Y_{0}]}{E[Y_{1}]} \times \frac{1/p_3}{T} \times \frac{A}{A} \times \frac{1/EDR^2}{1/MDD^2} $$

After simplification, we get:

$$ \frac{\hat{N}^{(PIX)}}{\hat{N}^{(PIF)}} = \frac{E[Y_{0}]}{E[Y_{1}]} \times \frac{1}{p_3T} \times \frac{MDD^2}{EDR^2} \times \frac{1}{H} $$

For simplicity, we will refer to these components as additive log ratios, $\Delta$.
$\Delta_{OBS}$ is the observed log ratio of the two estimators:
$\Delta_{EXP}$ is the expected log ratio based on the components.
$\Delta_{OBS} = \Delta_{EXP} + \epsilon$, where $\epsilon$ is the residual
not accounted for by any of the $\Delta$ components:

$$ \Delta_{OBS} = \Delta_{R} + \Delta_{T} + \Delta_{D} + \Delta_{H} + \epsilon = \Delta_{EXP} + \epsilon$$

$\Delta_{R}$ is the component related to roadside bias in counts,
$\Delta_{T}$ is the component related to time adjustment,
$\Delta_{A}$ is the component related to sampling area adjustments,
$\Delta_{H}$ is the component related to habitat sampling bias.

### PIF population size estimates

We used the 2013 version of the data set, filtered for BCR 6 Alberta.

### Model based population size estimates

We followed the methodology outlined in Ball et al. 2016 and replicated the 
analysis and prediction for `r nrow(pop)` species presented here.
To apply the methodology more generally across species, we did not include
terrain metrics as covariates, and we did not identified high suitability
areas to be considered at larger spatial scales for each individual species.
We only used amount of human footprint in 1 km$^2$ buffer areas.

```{r map,echo=FALSE,dev='png',fig.height=10,fig.width=5}
op <- par(mar=c(1,1,1,1))
plot(hill, col=grey(0:100/100), legend=FALSE, bty="n",
    box=FALSE, axes=FALSE)
plot(r, legend=FALSE, col=topo.colors(50, alpha=0.35)[26:50], add=TRUE)
plot(BCR2AB, col=c("#00000060", NA, rep("#00000060", 11)), add=TRUE)
plot(xy[xy@data$PCODE!="BBSAB" & !(!is.na(xy2BCR) & xy2BCR==24),], add=TRUE, pch=19, col="white", cex=0.5)
plot(xy[xy@data$PCODE=="BBSAB" & !(!is.na(xy2BCR) & xy2BCR==24),], add=TRUE, pch=19, col="lightblue", cex=0.5)
plot(xy[xy@data$PCODE!="BBSAB" & !is.na(xy2BCR) & xy2BCR==24,], add=TRUE, pch=19, cex=0.5)
plot(xy[xy@data$PCODE=="BBSAB" & !is.na(xy2BCR) & xy2BCR==24,], add=TRUE, pch=19, col=4, cex=0.5)
legend("bottomleft", title="Surveys", pch=c(19, 19, 19, 21), bty="n",
    col=c("blue", "lightblue", "black", "black"),
    legend=c("BBS in BCR 6", "BBS outside", "Off road in BCR 6", "Off road outside"))
par(op)
```


### Roadside count bias

We estimated expected density values for roadside and off-road habitats
given tha sample hased habitat composition along the roadside samples.
The ratio of expected counts is the same as the ratio of the densities,
thus we calculated the ratio of mean densities for this component.

### Time adjustment

We averaged the removal model based probabilities for 3-minute counts
over the surveys used for each species.

### Sampling area adjustment

We averaged the distance sampling based EDR values 
over the surveys used for each species.

### Habitat sampling bias

We calculated the proportion of the landbase ($a_k$, $k=1,\ldots,K$) in $K=19$ 
habitat classes. We also calculated proportion of BBS stops ($w_k$) 
clipped to BCR 6 within each of these $K$ classes.

A sample based estimator of total abundance is $N=A \sum_{k=1}^{K} D_k w_k$, 
where $A$ is the area of the region. If sampling is proportional to availability
($w_k = a_k$), the estimator is unbiased. Otherwise, the sampling
bias is characterized by the ratio 
$H = (\sum_{k=1}^{K} D_k w_k) / (\sum_{k=1}^{K} D_k a_k)$.

Here are the proportions for the whole extent of BCR 6 in Alberta (`a`) and
the BBS stops within (`w`):

```{r tab,echo=FALSE}
tmp <- 100*AWhab
colnames(tmp) <- c("a", "w")
knitr::kable(tmp, digits=2)
```

## Results

We found that pixel based population size estimates were on average
`r round(mean(pop$Npix/pop$Npif), 2)` (min.: `r round(min(pop$Npix/pop$Npif), 2)`, 
max: `r round(max(pop$Npix/pop$Npif), 2)`) times higher than PIF estimates.
Population size estimates are provided in Appendix.

```{r estimates,echo=FALSE}
dots_box_plot <- function(mat, ...) {
    set.seed(1)
    rnd <- runif(nrow(mat), -0.1, 0.1)
    boxplot(mat, range=0, ...)
#    for (i in 2:ncol(mat))
#        segments(x0=i+rnd-1, x1=i+rnd, y0=mat[,i-1], y1=mat[,i], col="lightgrey")
    for (i in 1:ncol(mat))
        points(i+rnd, mat[,i], pch=19, col="#00000080")
    boxplot(mat, range=0, add=TRUE, col="#ff000020", names=NA)
    invisible(NULL)
}

par(las=1)
dots_box_plot(pop[,c("Npif", "Npix")], ylab=expression(hat(N)), names=c("PIF", "PIX"))
```

The $\Delta$ components are shown in the next figure. The roadside count 
($\Delta_R$) and the habitat sampling component  ($\Delta_H$) had
means close to 0 indicating a wide range of responses from
individual species but no systematic bias.
As opposed to this, the time  ($\Delta_T$) and area  ($\Delta_A$)
adjustments presented a systematic bias indicating that these adjustments would
drive pixel based population size estimates higher than PIF estimates.

```{r deltas,echo=FALSE}

mat <- pop[,c("DeltaObs", "DeltaExp", "DeltaR", "DeltaT", "DeltaA", "DeltaH")]
colnames(mat) <- c("OBS", "EXP", "R", "T", "A", "H")
par(las=1)
dots_box_plot(mat, ylab=expression(Delta))
abline(h=0, col=1, lwd=1,lty=2)
```

We used linear regression, to see if the $\Delta$ components capture the
whole relationship between the two sets of population size estimates.
We expected the intercept to be close to 0, while the other coefficients to 
be around 1.

```{r model,echo=FALSE}
mod <- lm(DeltaObs ~ DeltaR + DeltaT + DeltaA + DeltaH, pop)
summary(mod)
```

As expected, the intercept was close to zero, and effect sizes close to 1.
The largest deviation was detected for the sampling area ($\Delta_A$) component.
The residual variance was `r round(summary(mod)$sigma^2, 3)`,
and the model explained `r round(100*summary(mod)$r.squared, 1)` % of the total 
variation. The percent variation explained by the individual
components is shown in the following table:

```{r anova,echo=FALSE}
an <- anova(mod)
an$Percent <- 100 * an[["Sum Sq"]] / sum(an[["Sum Sq"]])
an <- an[c("Df", "Sum Sq", "Percent", "Mean Sq", "F value", "Pr(>F)")]
an
```

The roadside count component ($\Delta_R$) explained the largest portion of the variation,
followed by the habitat sampling component ($\Delta_H$),
the area adjustment ($\Delta_A$),
and the time adjustment component ($\Delta_T$).

## Summary

The following main points were identified:

1. Pixel based and PIF population size estimates are (very) different.
2. Roadside count and habitat sampling biases vary greatly across species and these are more important in terms in explaining variation in estimates across species.
2. The effects area and time adjustments were systematic and more uniform across species explaining the average magnitude of differences among the two population size estimators.

Implications:

1. Roadside related biases need to be better understood, which is complicated by the fact that species responses are very different.
2. Habitat sampling bias can lead to under- and over-sampling when using a sample based estimator without accounting for sampling bias. A stratified population size estimator can potentially account for this bias. (This could have implications for trend estimation based on BBS.)
3. Better quantification of time and area adjustments can be easily incorporated into current PIF approach.

## Appendix

Population size estimates, million singing individuals. Table is ordered by
pixel based estimates.

```{r appendix,echo=FALSE}
tmp <- pop[,c("English_Name", "Npif", "Npix", "NpixLo", "NpixHi", "DeltaObs")]
tmp <- tmp[order(rownames(tmp)),]
knitr::kable(tmp, digits=2)
```

