---
title: "Avian population size estimates in BCR 6 Alberta"
author: "Peter Solymos (solymos@ualberta.ca)"
output:
  pdf_document: default
  word_document: default
---

# Introduction

Our estimates of current Boreal population sizes in Alberta ($\hat{N}^{(QPAD)}$)
are based on an exhaustive model based prediction approach
using province wide vegetation and footprint maps.
Existing populations size estimates provided by
Partners in Flight (PIF) provide another set of
population size estimates ($\hat{N}^{(PIF)}$; we used numbers
referring to singing individuals, i.e. dividing the 2013
PIF estimates by the pair adjustment) for the same species
within Birds Conservation Region 6 in Alberta.

PIF used a population-size estimator based on
species-specific counts from the roadside 
North American Breeding Bird Survey (BBS). 
PIF organized a peer review of the PIF approach done by Thogmartin (2006)
to address concerns by the ornithological community.
The criticism identified the following main assumptions that might 
bias populations size estimates by PIF:

* _Time adjustment_: factors were derived based on average counts as a function of
  time of day and time of year to come up with a factor
  for each species ($T$) that would adjust mean counts relative to the
  yearly and daily maximum for counts.
* _Maximum detection distance_: the detection-distance adjustment 
  defines the effective area surveyed based on a species-specific 
  maximum detection distance ($MDD$) that birds can be heard or seen by observers.
* _Roadside counts_: roadside-survey counts are equivalent to survey counts in off-road areas.
* _Habitat sampling_: BBS routes sample avian habitats in proportion to their availability.

There is evidence that some of these assumptions are violated:

* _Time adjustment_: Solymos et al. in preparation (duration ms.).
* _Maximum detection distance_: Matsuoka et al. 2012.
* _Roadside counts_: many papers.
* _Habitat sampling_: Matsuoka et al. 2011, Van Wilgenburg et al. 2015.

Our procedure, in contrast to the PIF approach, makes different assumptions
and tries to improve the estimators:

* _Time adjustment_: we have used offsets based on probabilities
  derived from removal-sampling to take into account availability bias.
* _Maximum detection distance: we have used effective detection
  distances ($EDR$) 
  derived from distance-sampling to take into account detectability bias.
* _Roadside counts_: we combined roadside and off-road surveys
  used a model based correction to account for the roadside bias.
* _Habitat sampling_: we have used a wider set of surveys
  in Alberta (roadside and off-road surveys) and applied
  a model based approach (whereas the PIF approach was a design based
  approach).

We compared the 2 sets of population size estimates for 97 bird species
in BCR 6 Alberta. We partition the differences into subsets to see
how the different assumptions affect the estimates.

# Methods

The ratio of the population estimates derived from our approach and the PIF
approach gives us the \emph{observed bias}: 
$\Delta^{(OBS)} = \hat{N}^{(QPAD)} / \hat{N}^{(PIF)}$ (the term
'bias' is used here without implying that any of these numbers are
the truth).
We expected that the magnitude and direction of the observed bias was related 
to the different assumptions. To evaluate this, we
compared the estimators. 

The QPAD estimator is defined as:
$\hat{N}^{(QPAD)} = A E[Y] (1/p_{3}) (1/\pi EDR^2)$.
The PIF estimator is defined as:
$\hat{N}^{(PIF)} = A E[Y^{(ROAD)}] T (1/\pi MDD^2)$.
We used the inverse of the probability of singing within a 3-minutes
survey interval ($p_{3}$) as an equivalent to the time adjustment
because BBS surveys are 3-minutes unlimited distance counts, thus
the correction factor for those would be defined as $1/p_{3}$.

It follows that the \emph{expected bias} is
$\Delta^{(EXP)} = (\frac{Y}{Y^{(ROAD)}}) (\frac{1/p_{3}}{T}) (\frac{\pi MDD^2}{\pi EDR^2})$
(note that area of the study region $A$ cancels out).
For notational simplicity, we refer to the components of the bias as:
$\Delta^{(EXP)} = \delta^{(ROAD)} \delta^{(TIME)} \delta^{(DIST)}$.
The time and distance components of the bias were calculated from
our estimates. The road related component was estimated
by the road effect from our modeling, as that represents
the effect of road relative to an off-road point location 
(i.e. without the habitat sampling bias).

We further assumed that the observed and expected bias will differ, 
thus we introduced the residual bias as: 
$\Delta^{(OBS)} = \Delta^{(EXP)} \delta^{(RESID)}$.
Bias is 1 when QPAD and PIF components equal. A larger than 1 value indicate that
the QPAD component is larger than the PIF component.

# Results

```{r echo=FALSE,message=FALSE,results='hide',warning=FALSE}
library(knitr)

bounded_density <- function(x, interval=c(-Inf, Inf), ...) {
    require(intrval)
    interval <- sort(interval[1L:2L])
    if (!all(x %()% interval))
        stop("found values of x outside of open interval")
    a <- interval[1L]
    b <- interval[2L]
    bounds  <- is.finite(interval)
    if (!bounds[1L] && !bounds[2L]) { # (Inf, Inf)
        f <- finv <- function(x) x
    }
    if (bounds[1L] && !bounds[2L]) { # (a, Inf)
        f <- function(x)
            log(x-a)
        finv <- function(z)
            exp(z) + a
    }
    if (!bounds[1L] && bounds[2L]) { # (Inf, b)
        f <- function(x)
            log(b-x)
        finv <- function(z)
            b-exp(z)
    }
    if (bounds[1L] && bounds[2L]) { # (a, b)
        f <- function(x)
            qlogis((x-a) / (b-a))
        finv <- function(z)
            plogis(z) * (b-a) + a
    }
    fx <- f(x)
    d <- density(fx, ...)
    v <- d$x
    dv <- diff(v)
    h <- d$y
    n <- length(h)
    dh <- rowMeans(cbind(h[-n], h[-1L]))
    A <- dv * dh
    vinv <- finv(v)
    hinv <- A / abs(diff(vinv))
    hinv <- c(hinv[1L], hinv)
    data.frame(x=vinv, y=hinv)
}

if (FALSE) {
x <- rnorm(100)

op <- par(mfrow=c(2,2))
range(x)
d1 <- bounded_density(exp(x))
d2 <- bounded_density(exp(x), c(0, Inf))
plot(bounded_density(x, type="l"))

range(exp(x))
d1 <- bounded_density(exp(x))
d2 <- bounded_density(exp(x), c(0, Inf))
plot(d1, type="l", ylim=c(range(d1$y, d2$y)))
lines(d2, col=2)

range(plogis(x)*4-1)
d1 <- bounded_density(plogis(x)*4-1)
d2 <- bounded_density(plogis(x)*4-1, c(-1, 3))
plot(d1, type="l", ylim=c(range(d1$y, d2$y)))
lines(d2, col=2)

range(-exp(x)+3)
d1 <- bounded_density(-exp(x)+3)
d2 <- bounded_density(-exp(x)+3, c(-Inf, 3))
plot(d1, type="l", ylim=c(range(d1$y, d2$y)))
lines(d2, col=2)
par(op)
}

cp1 <- function(x, p=c(0, 1), nmin=5, interval=c(-Inf, Inf), ...) {
    x2 <- x[!is.na(x)]
    q <- quantile(x2, p)
    x2 <- x2[x2 > q[1] & x2 < q[2]]
    if (length(unique(x2)) < nmin) {
        m <- mean(x2)
        s <- sd(x2)
        v <- seq(min(min(x2), m-4*s), max(max(x2), m+4*s), length.out = 512)
        v <- v %()% interval
        d <- dnorm(v, m, s)
        v <- c(v[1L], v, v[length(v)])
        d <- c(0, d, 0)
        out <- data.frame(h=v, w=d/max(d))
    } else {
        d <- bounded_density(x2, interval=interval, ...)
        out <- data.frame(h=d$x, w=d$y/max(d$y))
    }
    out
}
carrot <- function(x, p=c(0,1),
col="#FF664D80", border="#FF664D",
#col="#40E0D180", border="#40E0D1",
ylim, ylab="", xlab="", main="", nmin=5, interval=c(-Inf, Inf), ...) {
    xx <- apply(x, 2, cp1, p=p, nmin=nmin, interval=interval, ...)
    if (missing(ylim))
        ylim <- range(x, na.rm=TRUE)
    plot(0, type="n", axes=FALSE, ann=FALSE, xlim=c(0.5, ncol(x)+0.5),
        ylim=ylim)
    axis(1, 1:ncol(x), colnames(x), lwd=0)
    axis(2)
    title(ylab=ylab, xlab=xlab, main=main)
    for (i in 1:ncol(x)) {
        if (!is.null(xx[[i]])) {
            polygon(0.45*c(-xx[[i]]$w, rev(xx[[i]]$w))+i,
                c(xx[[i]]$h, rev(xx[[i]]$h)), col=col, border=border)
            #lines(c(i,i), range(xx[[i]]$h), col=col, lwd=2)
        }
        lines(c(i-0.2, i+0.2), rep(median(x[,i], na.rm=TRUE), 2), lwd=3)
    }
    #points(1:ncol(x), colMeans(x, na.rm=TRUE), pch=21, cex=1.5)
    invisible(NULL)
}
pop <- read.csv("~/Dropbox/bam/PIF-AB/qpad-pif-results.csv")
rownames(pop) <- pop$Species_ID
rnd <- runif(nrow(pop), -0.1, 0.1)
```

Population sizes were quite different.

```{r pop-size,echo=FALSE,fig.width=8,fig.height=6}
par(las=1)
boxplot(pop[,c("Npif", "Nqpad")], col=NA, range=0, ylab="M singing inds.")
for (i in 1:nrow(pop))
    lines(c(1,2), pop[i,c("Npif", "Nqpad")], col="#00000020")
points(1+rnd, pop[,"Npif"], col="#00000040", pch=19)
points(2+rnd, pop[,"Nqpad"], col="#00000040", pch=19)
boxplot(pop[,c("Npif", "Nqpad")], range=0, add=TRUE, col=NA)
```

The median observed bias between QPAD and PIF population estimates
was `r round(median(pop$DeltaObs), 2)` and ranged between `r round(min(pop$DeltaObs), 2)` and 
`r round(max(pop$DeltaObs), 2)`.

```{r echo=FALSE,fig.width=8,fig.height=6}
par(las=1)
mat <- log(pop[,c("DeltaObs", "DeltaExp", "DeltaRoad", "DeltaTime", "DeltaDist", "DeltaRes")])
if (FALSE) {
boxplot(mat, range=0, col=NA, ylab="bias")
for (i in 2:ncol(mat))
    segments(x0=i+rnd-1, x1=i+rnd, y0=mat[,i-1], y1=mat[,i], col="#00000020")
for (i in 1:ncol(mat))
    points(i+rnd, mat[,i], col="#00000040", pch=19)
abline(h=0, col=2, lwd=2)
boxplot(mat, range=0, add=TRUE, col=NA)
}
carrot(mat, ylab="bias")
for (i in 2:ncol(mat))
    segments(x0=i-1, x1=i, y0=mat[,i-1], y1=mat[,i], col="#00000020")
abline(h=0, col=1, lwd=1)
```

The components of the bias showed interesting trends:

* the road related bias and the residual bias showed large
  variation across the species with a close to 0 median value,
* the time related bias was on average positive, but relatively small,
  compared to the other components,
* the distance related bias was consistently higher for all species, 
  with magnitudes between `r round(min(pop$DeltaDist), 2)` and `r round(max(pop$DeltaDist), 2)`
  which is consistent with the expected distance related biases
  reported by Matsuoka et al. (2012).

The median of the expected bias across the 97 species was `r median(pop$DeltaExp)`,
which is very different from the observed bias, thus the residual bias in
considerable (and negative on the log scale, compensating for the
large positive expected bias on the log scale).

The time, distance and road components of the bias were quantified
by estimated model parameters and published values.
The residual bias is the only piece that was not quantified.
Thus, the residual bias must be related to the last assumption,
namely that BBS routes sample avian habitats in proportion to their availability.

We expected large residual bias for species whose preferred habitats
are under-sampled by roads.
To express the relationship between roads and preferred habitats
of the species, we calculated the road avoidance index:
$RAI = 1-\sum^{H}_{h=1} R_{h} O_{h}$, a covariance type measure
where $O_{i}$ is proportion of predicted population expected in habitat
class $h$ (based on predicted abundance distributed among major land cover types).
$R_{h}$ was the proportion of roadside samples within habitat class $h$.
The product indicates the joint occurrence of a road and the species,
thus the one-complement of the index refers to road avoidance.

We calculated $RAI$ for road as well (as if it was a species), and the
value was `r round(pop$RAIroad[1], 2)`. It was used to center the $RAI$ values
so that species with negative values prefer roads dispropotionately,
whereas species with positive values are avoiding it.

We found high agreement between the value of the
log residual bias and road avoidance index.
Residual bias was less than one for species with small
RAI values, whereas the residual bias was greater than one 
for species with RAI close to 1 (high avoidance).

```{r echo=FALSE,fig.width=8,fig.height=6}
par(las=1)
with(pop, plot(RAIc, log(DeltaRes), type="n"))
abline(h=0, v=0, col=1, lty=2)
abline(lm(log(DeltaRes) ~ RAIc, pop), col=1)
with(pop, text(RAIc, log(DeltaRes), rownames(pop), cex=0.75, col=4))
```


# Appendix

Million singing individuals of the different species estimated in BCR 6 Alberta.

```{r echo=FALSE}
kable(pop[,c("English_Name", "Npif", "Nqpad")])
```


