---
title: "Standardized Species Output Metadata (v2018)"
author: "Peter Solymos, <solymos@ualberta.ca>"
date: "`r as.Date(Sys.time())`"
output: word_document
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
#options(scipen = 999)
HEAD <- c("Header", "Type", "Required", "Description")
ALIGN <- c("lccl")
```

## Introduction

The goal of this document to describe species output standards to automate the creation of

1. figures for development website (SC review and soliciting feedback from experts before releasing new version),
2. make custom predictions (given same backfilled version) for various purposes (BMF, scenario based management)

This would greatly reduce 'project overhead' that is
usually significant time spent on not creating additional value 
(handing off between SC members and between SC and IC).

## Versions

Version of 'backfilled' soil/vegetation layer used alongside of the verified footpring info for modelling is noted for each taxa in the final
compendium of results (within the **cure4insect** package).
Version are denoted by the year of summer surveys when most of the data
was collected for the update. I.e. v2018 in this document.
Updates might not be backwards compatible bue to changes in land cover classification or the soil/vegetation layer.

## Format

Because all of us running the models use R, it seems logical to use native R format (.RData files)
for sharing the results. This can be further processed and written into other 
formats upon request, e.g. CSV, Excel, GeoTIFF, etc.

I am proposing the following structure:

**Files** --- Each taxon have its own file, file name reflecting the taxon and date of creation (using ISO standard YYYY-MM-DD format), e.g. `ABMI-vplant-results-2019-03-01.RData`. A taxon can be divided into more than one file, e.g. native vs. non-native species, or snow-tracking vs. camera -- which also needs to be reflected in the file name.

**Objects** --- In each taxon's file, we have a standard set of objects 
(usually tables, like matrices or data frames) as listed in the next section.
These tables list species as rows (Species ID as row name) and various fields as columns (standardized column names listed below). Cell values in these tables usually refer to 
parameter estimates, predictions, or summary statistics.

## Objects

Objects are 2D or 3D arrays. 

* The 1st dimension represent species, names along the 1st dimension are called species IDs.
* The 2nd dimension represent fields,  names along the 2nd dimension must match _all_ the fields listed below. Objects can have other fields, and ordering if these fields can be arbitrary.
* The 3rd dimension is optional, when it is present, it usually represents bootstrap replicates aling the 1st and 2nd dimensions (i.e. coefficient estimates coming from bootstrap runs), and the number of replicates (e.g. 100) defines the 3rd dimension size. Camera mammals have separate models for summer and winter, this can also be used as the 3rd dimension, adding distinguishing names along that dimension.


Field types are defined as the following:

* Chr: character or factor, or coercible to these.
* Bin: logical (`TRUE`/`FALSE`) or coercible to these (i.e. 0/1).
* Int: discrete numbers, usually non-negative
* Num: real valued numbers represented with 6 digits precision.

The following sections describe the schema for the individual objects.

Notes:

* When a given variable was not estimated for a species, put `NA` when the variable was not even considered, and put 0 when an estimates missing due to model selection (i.e. it was part of the original scope).
* When a coefficient refers to multiple land cover types (because those were pooled for modeling), repeat the same value under the multiple names.

### General

#### Species lookup table

One special table is the species lookup table, that lists all the species IDs 
that serve as row names for the other objects.
This table links together all the other objects through its Species ID field, 
which is used as row name in other objects (and this used as file name in the past).
Not required fields can be blank (for character/factor types)

* Object name: `Lookup`
* Object format: data frame (colnames as in table)

```{r}
Lookup0 <- matrix(byrow=TRUE, ncol=4, dimnames=list(NULL, HEAD), data=c(
"SpeciesID", "Chr", "Y", "Primary key for species.",
"ScientificName", "Chr", "Y", "Scientific name(s) of possible multiple taxa included, must match names from taxonomic workbench / Oracle.",
"TSNID", "Chr", "", "Use only of ScientificName has not been grouped/altered.",
"CommonName", "Chr", "", "Provide is it exists.",
"ModelNorth", "Bin", "Y", "Model based on north data exists and we decided to show it.",
"ModelSouth", "Bin", "Y", "Model based on south data exists and we decided to show it.",
"Nonnative", "Bin", "Y", "Non native status, assumed native if missing.",
"LinkHabitat", "Chr", "Y", "Link function used in north/south habitat models, i.e. logit, log, identity (usually the same for all spp except for habitat elements).",
"LinkSpclim", "Chr", "Y", "Link function used in north/south space/climate models, i.e. logit, log, identity (usually the same for all spp except for habitat elements).",
"Comments", "Chr", "", "Known issues, general disclaimers about secies or groups of species."
))
kable(Lookup0, row.names=FALSE, align=ALIGN, padding = 2)
```

#### Observations

Have the species data used for modeling in a table (matrix or data frame) 
with samples as rows and Species IDs as column names.
Cell values are the observations (detections, counts, densities).
Zero (0) indicates that the species at that site or in that sample was not 
detected.
Additional columns can indicate if any non-ABMI data were used, so that
we can report on these separately.
Add longitude and latitude, so that these extra locations can be mapped.
Use public coordinates for ABMI sites,
and provice public coordinates for nearest ABMI site in case the location
cannot be shared.


### South

#### Use/availability

Summary statistics of the datactions and land cover composition.

* Object name: `UseavailSouth`
* Object format: matrix or data frame (rownames are SpeciesID's, colnames as in table)


```{r}
UseavailSouth0 <- data.frame(c(
    "Productive",
    "Clay",
    "Saline",
    "RapidDrain",
    "Crop",
    "TameP",
    "RoughP",
    "UrbInd",
    "HardLin",
    "SoftLin"), 
    "Num", "Y", "Use availability statistic.")
colnames(UseavailSouth0) <- HEAD
kable(UseavailSouth0, row.names=FALSE, align=ALIGN)
```


#### Soil and HF coefficients

Backtransformed (i.e. not log/logit scale) relative abundances from the habitat part of the models.
Note: this table only covers land cover types and not spatial covariates used in the habitat modeling (e.g. pAspen).

* Object name: `CoefSouth`
* Object format: matrix, a data frame (rownames are SpeciesID's, colnames as in table), or a 3D array


```{r}
CoefSouth0 <- data.frame(c(
    "Productive",
    "Clay",
    "Saline",
    "RapidDrain",
    "Water",
    "Crop",
    "TameP",
    "RoughP",
    "UrbInd",
    "HardLin",
    "SoftLin"), 
    "Num", "Y", "Relative abundance estimate.")
colnames(CoefSouth0) <- HEAD
kable(CoefSouth0, row.names=FALSE, align=ALIGN)
```

Derivative values:

* AverageCoef is defined as arithmetic mean of non HF classes (Productive, Clay, Saline, RapidDrain).
* SoftLin10: defined as 0.9 x AverageCoef + 0.1 x SoftLin.
* HardLin10: defined as 0.9 x AverageCoef + 0.1 x HardLin.

Uncertainties: two additional tables can be added. The table has same rows/columns as `CoefSouth`. These are the error bars that would be 
included on the coefficient figures.

* `LowerSouth`: lower limits of the 90% confidence intervals, pre calculated
* `UpperSouth`: upper limits of the 90% confidence intervals, pre calculated

As an alternative, `CoefSouth` can be a 3D array in which case intervals are based on quantiles. This is when the whole modeling is bootstrapped.

Optionally, a similarly structured table (`HonestSouth`) can be created to indicate which coefficients were estimeted togather. I.e. instead of coefficient values, one would use (1, 1, 2, 2) to indicate that (A, B, C, D) classes were estimates as (A+B, A+B, C+D, C+D). Either numeric or character representation would work.

#### Space and climate coefficients

Untransformed (i.e. on log/logit scale) parameter estimates from the space climate part of the models.
Note: this table contains pAspen that was estimated in the habitat part of the model, but we consider it part of this table.

* Object name: `SpclimSouth`
* Object format: matrix, a data frame (rownames are SpeciesID's, colnames as in table), or a 3D array

```{r}
SpclimSouth0 <- data.frame(c(
    "pAspen",
    "Intercept",
    "Lat",
    "Long",
    "AHM",
    "PET",
    "FFP",
    "MAP",
    "MAT",
    "MCMT",
    "MWMT",
    "Lat2",
    "Long2",
    "LatLong",
    "MAPPET",
    "MATAHM",
    "MAPFFP",
    "MAT2",
    "MWMT2"),
    "Num", "Y", "Parameter estimates.")
colnames(SpclimSouth0) <- HEAD
kable(SpclimSouth0, row.names=FALSE, align=ALIGN)
```

Variables are assumed untransformed, other than 2 indicating squared, and 2 names side-by-side indicating an interaction (product).

If a 3D array, 3rd dimension represent independent bootstrap estimates.

#### Auxiliary variables

Auxiliary variables (methodology related variables, snow days, etc.)
can be store in additional objects (i.e. `ExtrasNorth`)
with SPecies IDs as rownames and columns referring to these
additional parameters.

### North

#### Use/availability

Summary statistics of the datactions and land cover composition.

* Object name: `UseavailNorth`
* Object format: matrix or data frame (rownames are SpeciesID's, colnames as in table)


```{r}
UseavailNorth0 <- data.frame(c(
    "Deciduous",
    "Mixedwood",
    "WhiteSpruce",
    "Pine",
    "BlackSpruce",
    "TreedFen",
    "Open",
    "Wetland",
    "Water",
    "HFor",
    "Crop",
    "TameP",
    "RoughP",
    "UrbInd",
    "HardLin",
    "SoftLin"),
    "Num", "Y", "Use availability statistic.")
colnames(UseavailNorth0) <- HEAD
kable(UseavailNorth0, row.names=FALSE, align=ALIGN)
```

Notes: Open includes Bare.

#### Vegetation and HF coefficients

Backtransformed (i.e. not log/logit scale) relative abundances from the habitat part of the models.

* Object name: `CoefNorth`
* Object format: matrix, a data frame (rownames are SpeciesID's, colnames as in table), or a 3D array.


```{r}
CoefNorth0 <- data.frame(c(
    "WhiteSpruce_0-10",
    "WhiteSpruce_10-20",
    "WhiteSpruce_20-40",
    "WhiteSpruce_40-60",
    "WhiteSpruce_60-80",
    "WhiteSpruce_80-100",
    "WhiteSpruce_100-120",
    "WhiteSpruce_120-140",
    "WhiteSpruce_140+",
    "Pine_0-10",
    "Pine_10-20",
    "Pine_20-40",
    "Pine_40-60",
    "Pine_60-80",
    "Pine_80-100",
    "Pine_100-120",
    "Pine_120-140",
    "Pine_140+",
    "Deciduous_0-10",
    "Deciduous_10-20",
    "Deciduous_20-40",
    "Deciduous_40-60",
    "Deciduous_60-80",
    "Deciduous_80-100",
    "Deciduous_100-120",
    "Deciduous_120-140",
    "Deciduous_140+",
    "Mixedwood_0-10",
    "Mixedwood_10-20",
    "Mixedwood_20-40",
    "Mixedwood_40-60",
    "Mixedwood_60-80",
    "Mixedwood_80-100",
    "Mixedwood_100-120",
    "Mixedwood_120-140",
    "Mixedwood_140+",
    "BlackSpruce_0-10",
    "BlackSpruce_10-20",
    "BlackSpruce_20-40",
    "BlackSpruce_40-60",
    "BlackSpruce_60-80",
    "BlackSpruce_80-100",
    "BlackSpruce_100-120",
    "BlackSpruce_120-140",
    "BlackSpruce_140+",
    "TreedFen",
    "Shrub",
    "Grass",
    "Bare",
    "Water",
    "TreeShrubSwamp",
    "NonTreeFenMarsh",
    "CCWhiteSpruce_0-10",
    "CCWhiteSpruce_10-20",
    "CCWhiteSpruce_20-40",
    "CCWhiteSpruce_40-60",
    "CCWhiteSpruce_60-80",
    "CCPine_0-10",
    "CCPine_10-20",
    "CCPine_20-40",
    "CCPine_40-60",
    "CCPine_60-80",
    "CCDeciduous_0-10",
    "CCDeciduous_10-20",
    "CCDeciduous_20-40",
    "CCDeciduous_40-60",
    "CCDeciduous_60-80",
    "CCMixedwood_0-10",
    "CCMixedwood_10-20",
    "CCMixedwood_20-40",
    "CCMixedwood_40-60",
    "CCMixedwood_60-80",
    "Crop",
    "TameP",
    "RoughP",
    "UrbInd",
    "SoftLin",
    "HardLin"), 
    "Num", "Y", "Relative abundance estimate.")
colnames(CoefNorth0) <- HEAD
kable(CoefNorth0, row.names=FALSE, align=ALIGN)
```

Derivative values:

* AverageCoef is defined as arithmetic mean of non HF classes (not CC, Crop, TameP, RoughP, UrbInd, SoftLin, HardLin).
* SoftLin10: defined as 0.9 x AverageCoef + 0.1 x SoftLin.
* HardLin10: defined as 0.9 x AverageCoef + 0.1 x HardLin.

Uncertainties: two additional tables can be added. The table has same rows/columns as `CoefNorth`. These are the error bars that would be 
included on the coefficient figures.

* `LowerNorth`: lower limits of the 90% confidence intervals, pre calculated
* `UpperNorth`: upper limits of the 90% confidence intervals, pre calculated

As an alternative, `CoefNorth` can be a 3D array in which case intervals are based on quantiles. This is when the whole modeling is bootstrapped.

Optionally, a similarly structured table (`HonestNorth`) can be created to indicate which coefficients were estimeted togather. I.e. instead of coefficient values, one would use (1, 1, 2, 2) to indicate that (A, B, C, D) classes were estimates as (A+B, A+B, C+D, C+D). Either numeric or character representation would work.

#### Space and climate coefficients

Untransformed (i.e. on log/logit scale) parameter estimates from the space climate part of the models.

* Object name: `SpclimNorth`
* Object format: matrix, a data frame (rownames are SpeciesID's, colnames as in table), or a 3D array

```{r}
SpclimNorth0 <- data.frame(c(
    "pAspen",
    "Intercept",
    "Lat",
    "Long",
    "AHM",
    "PET",
    "FFP",
    "MAP",
    "MAT",
    "MCMT",
    "MWMT",
    "Lat2",
    "Long2",
    "LatLong",
    "MAPPET",
    "MATAHM",
    "MAPFFP",
    "MAT2",
    "MWMT2"),
    "Num", "Y", "Parameter estimates.")
colnames(SpclimNorth0) <- HEAD
kable(SpclimNorth0, row.names=FALSE, align=ALIGN)
```

Variables are assumed untransformed, other than 2 indicating squared, and 2 names side-by-side indicating an interaction (product).

If a 3D array, 3rd dimension represent independent bootstrap estimates.

```{r eval=FALSE}
Lookup <- data.frame(
    SpeciesID=c("Spp1", "Spp2", "Spp3", "Spp4"),
    ScientificName=c("Homo sapiens", "Tirannosaurus rex", "Paramecium spp.", "Felis catus"),
    TSNID=c("TSN000", "", "ABMI666", "TSN999"),
    CommonName=c("Human", NA, NA, NA),
    ModelNorth=c(TRUE, FALSE, FALSE, TRUE),
    ModelSouth=c(TRUE, TRUE, FALSE, FALSE),
    SizeNorth=c(100, NA, NA, 10),
    SizeSouth=c(200, 1, NA, NA),
    Nonnative=c(TRUE, FALSE, FALSE, TRUE),
    Link=c("identity", "logit", "logit", "log"),
    Comments=c("", "Extinct.", "Oh crap, what is this !?", "Meow"))
rownames(Lookup) <- Lookup$SpeciesID
f <- function(x) {
    matrix(rnorm(nrow(x) * nrow(Lookup)), 
        nrow(Lookup), nrow(x),
        dimnames=list(rownames(Lookup), x[,1]))
}
CoefNorth <- exp(f(CoefNorth0))
CoefSouth <- exp(f(CoefSouth0))
SpclimNorth <- f(SpclimNorth0)
SpclimSouth <- f(SpclimSouth0)
UseavailNorth <- f(UseavailNorth0)
UseavailSouth <- f(UseavailSouth0)
save(list=c("CoefNorth", "CoefSouth",
    "Lookup", "SpclimNorth", "SpclimSouth", 
    "UseavailNorth", "UseavailSouth"), file="StandardizedOutput.RData")
```

