---
title: 'Assessing the predictive capabilities of songbird abundance models for cumulative effects management in Alberta - Appendix 3'
author: "Péter Sólymos, Lionel Leston, Judith D. Toms, and Erin M. Bayne"
date: "Version: 1.0"
output:
  pdf_document:
    toc: false
    toc_depth: 3
    number_sections: false
documentclass: report
fontsize: 10pt
classoption: letterpaper
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
---

```{r preamble,include=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.pos="H", fig.align="center")
SPP <- c("ALFL", "AMCR", "AMRE", "AMRO", "BAWW", "BBWA", "BCCH", "BHVI", 
"BOCH", "BRCR", "CCSP", "CEDW", "CHSP", "CONW", "CORA", "COYE", 
"DEJU", "GCKI", "GRAJ", "HETH", "LCSP", "LEFL", "LISP", "MAWA", 
"MOWA", "OVEN", "PAWA", "PISI", "RBGR", "RBNU", "RCKI", "REVI", 
"RWBL", "SOSP", "SWSP", "SWTH", "TEWA", "TRES", "WAVI", "WETA", 
"WIWR", "WTSP", "WWCR", "YBSA", "YEWA", "YRWA")
ROOT <- "/Users/Peter/GoogleWork/abmi/validation"
tax <- read.csv("~/repos/abmispecies/_data/birds.csv")
rownames(tax) <- tax$AOU
tax <- tax[SPP,]
```

This appendix present species specific results for `r length(SPP)`
songbird species. For each species we present two visual summary plots.
Both these plots are based on the most complex model (stage HF).

The first plot present the relationship between the observed counts
and the predicted abundances. The perfect alignment is through the
1:1 line (scattered line). Facets show the different spatial scales.
The fitted line (brown solid line) and confidence bands (in grey)
is based on linear regression. The Poisson distribution based
prediction intervals are displayed via the dot colors (green:
observed value is inside the 95% prediction interval; yellow:
observed value is outside the 95% prediction interval).

The second plot shows the results from the randomization.
The two validation statistics ($r$: rank correlation,
$R^2$ coefficient of determination; facet rows).
Results are displayed by randomization type (withing vs. between;
facet columns). Violin plots describe the randomization based null
distribution, colors and x-axis corresponds to spatial scales.
The black dot represents the non-randomized statistic values.


```{r results='asis'}
for (spp in SPP) {
  f1 <- file.path(ROOT, "species", paste0("comb-", spp, ".png"))
  #cat("\n\n\\newpage\n\n")
  cat("\n\n")
  cat("##", paste0(tax[spp, "species"], " (_", tax[spp, "scinam"], "_; ", spp, ")"))
  cat("\n\n")
  cat("![](", f1, ")", sep="")
}
```


