---
title: "GoF metrics"
author: "Peter Solymos"
date: "May 4, 2017"
output: pdf_document
---

## Deviance R-squared

* Null deviance: 2*(ll_saturated - ll_null)
* Null deviance df: df_sturated - df_null = n-1
* Residual deviance: 2*(ll_saturated - ll_full)
* Residual deviance df: df_sturated - df_full = n-(p+1)

Need to further adjust when null model includes more than the intercept.

p is the number of coefficients not including the intercept.

H0: the null model (no predictors other than a constant term) fits 
the data at least as well as our model.

H1: full model fits the data better than the null model.

```{r}
library(ResourceSelection)

fun_R2 <-
function(observed, fitted, distr=c("binomial", "poisson"),
size=1, null=NULL, p=0, alpha=0.05)
{
    distr <- match.arg(distr)
    if (is.null(null))
        null <- mean(observed)
    if (distr == "poisson") {
        ll0 <- sum(dpois(observed, null, log=TRUE))
        lls <- sum(dpois(observed, observed, log=TRUE))
        llf <- sum(dpois(observed, fitted, log=TRUE))
    } else {
        ll0 <- sum(dbinom(observed, size, null, log=TRUE))
        lls <- sum(dbinom(observed, size, observed/size, log=TRUE))
        llf <- sum(dbinom(observed, size, fitted, log=TRUE))
    }
    n <- length(observed)
    R2 <- 1 - (lls - llf) / (lls - ll0)
    R2adj <- 1 - (1 - R2) * ((n-1) / (n-(p+1)))
    D0 <- -2 * (ll0 - lls)
    DR <- -2 * (llf - lls)
    p_value <- 1 - pchisq(D0 - DR, p)
    list(R2=R2, R2adj=R2adj, Deviance=D0 - DR, Dev0=D0, DevR=DR, p=p,
        p_value=p_value)
}

y <- c(0, 1, 2, 0, 1, 2, 7, 3, 14, 4)
x <- 1:10
m <- glm(y ~ x, family=poisson)
fit <- fitted(m)
summary(m)
fun_R2(y, fit, "poisson", p=1)
anova(m, test="Chisq")
fun_R2(y, rev(fit), "poisson", p=1)

y <- c(0, 0, 1, 0, 1, 0, 1, 1, 1, 1)
x <- 1:10
m <- glm(y ~ x, family=binomial)
fit <- fitted(m)
summary(m)
fun_R2(y, fit, "binomial", p=1)
fun_R2(y, rev(fit), "binomial", p=1)

hoslem.test(y, fit) # p>0.05: no evidence for poor fit
hoslem.test(y, rev(fit)) # p<0.05: evidence for poor fit

```

## ROC and AUC

```{r}
library(pROC)
y <- c(0, 0, 1, 0, 1, 0, 1, 1, 1, 1)
x <- 1:10
m <- glm(y ~ x, family=binomial)
fit <- fitted(m)

ROC <- roc(y, fit)
ROC
plot(ROC)
## extract AUC without the fanfare
as.numeric(ROC$auc)
```

## Concordance

```{r}
library(epiR)
y <- rep(c(0, 0, 1, 0, 1, 0, 1, 1, 1, 1), 10)
x <- as.factor(rep(1:10, 10))
m <- glm(y ~ x, family=binomial)

B <- cbind(1:10, replicate(19, sample.int(10, 10, replace=TRUE)))
fit <- apply(B, 2, function(z) 
    fitted(glm(y ~ x, data=data.frame(y=y, x=x)[z,], family=binomial)))
epi.occc(fit)
```


## References

Hosmer D W, Lemeshow S 2000. Applied Logistic Regression. New York, USA: John Wiley and Sons.

Hastie, T. J. and Pregibon, D. (1992) Generalized linear models. Chapter 6 of Statistical Models in S eds J. M. Chambers and T. J. Hastie, Wadsworth & Brooks/Cole.

Xavier Robin, Natacha Turck, Alexandre Hainard, et al. (2011) pROC: an open-source package for R and S+ to analyze and compare ROC curves. BMC Bioinformatics, 7, 77. DOI: 10.1186/1471-2105-12-77.

Barnhart H X, Haber M, Song J (2002). Overall concordance correlation coefficient for evaluating agreement among multiple observers. Biometrics 58: 1020 - 1027.

Lin L (1989). A concordance correlation coefficient to evaluate reproducibility. Biometrics 45: 255 - 268.

Lin L (2000). A note on the concordance correlation coefficient. Biometrics 56: 324 - 325.


```{r eval=FALSE}
load("e:/peter/AB_data_v2017/misc/Deviance-test-data.RData")
rownames(d) <- d$SiteYear
all(rownames(Site.Prediction) == rownames(d))

obs <- as.matrix(d[,colnames(Site.Prediction)])
pred <- as.matrix(Site.Prediction)
Size <- d$TotalNoOfQU

i <- 1
fun_R2(obs[,i], pred[,i], distr="binomial", size=Size, p=1)
```
