---
title: "Modeling linear feature effects"
author: "P. Solymos"
date: '2015-07-15'
output:
  html_document:
    toc: no
  pdf_document:
    toc: no
    toc_depth: 2
---

# The model

$N = A_h D_h + A_l D_l$ where $h$ is surrounding habitat in a circular buffer
and $l$ is the linear feature, $A$ is area, $D$ is density, $N$ is abundance
at a point count.

$N = (1-p) A D_h + p A \delta D_h$, where $A=A_h + A_l$ and $D_l = \delta D_h$.
It follows that $D = N/A = D_h (1 - p + p \delta)$.

# The problem

We model $D_h = exp(X \beta)$, and $D = exp(X \beta) exp(\beta_l p)$.
But the scaling with $p$ is not the same in the intuitive model above,
and the model that is easy to implement in a `glm` framework. 
Let us plot $D/D_h$ in 2 different ways:

```{r}
delta <- 0.1
p <- seq(0, 1, by = 0.01)
op <- par(mfrow=c(1,3))
plot(p, exp(log(delta)*p), type = "l", col = 2, lwd = 2)
plot(p, 1-p+p*delta, type = "l", col = 2, lwd = 2)
plot(exp(log(delta)*p), 1-p+p*delta, type = "l", col = 2, lwd = 2)
abline(0, 1, lty = 2)
par(op)
```

# The solution

What would be a function of $p$ that would best approximate the intuitive model?

We will compare polynomial terms to estimate bias when the data is simulated
under the intuitive model.

```{r}
set.seed(1234)
n <- 1000
Dh <- 2
delta <- 0.1
p <- runif(n, 0, 1)
p <- p[order(p)]
lam <- (1-p)*Dh + p*delta*Dh
y <- rpois(n, lam)

summary(m <- glm(y ~ p, family=poisson))
exp(coef(m)[1]) # Dh
exp(coef(m)[1] + coef(m)[2]) # Dl
exp(coef(m)[2]) # delta

f <- fitted(m)
plot(p, lam, ylim = c(0, max(lam, f)))
lines(p, f)


summary(m <- glm(y ~ I(p^2), family=poisson))
exp(coef(m)[1]) # Dh
exp(coef(m)[1] + coef(m)[2]) # Dl
exp(coef(m)[2]) # delta

f <- fitted(m)
plot(p, lam, ylim = c(0, max(lam, f)))
lines(p, f)
```

The linear `glm` is not a bad approximation.


